---
title: "Carfilzomib_resistance"
author: "rbabaei"
date: "5/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, cache = T)
library(Biobase)
library(oligoClasses)

#Annotation
library(hgu133plus2.db)


#Quality control and pre-processing packages
library(oligo)
library(arrayQualityMetrics)

#Analysis and statistics packages
library(limma)
library(topGO)
library(ReactomePA)
library(clusterProfiler)

#Plotting and color options packages
library(ggplot2)
library(geneplotter)
library(RColorBrewer)
library(pheatmap)

#Formatting/documentation packages
library(dplyr)
library(tidyr)

#Helpers:
library(stringr)
library(matrixStats)
library(genefilter)
```

## Introduction
The aquired resistance of multiple myeloma model cell lines (KMS11 and KMS34) to anti cancer agent Carfilzomib has been studied by Riz I et.al. The article has been published in Oncotarget (PMID: 26109433). The KMS11 and KMS34 cells have been cultured in the presence of Carfilzomib and their gene expression profiles have been compared to the parental cells. Here we will analyze the microarray data of this study, which is accessible on GEO data set under GDS5826.


## Data importing and tidying
The CEL format of data is downloaded from GEO. First we are going to load the data into console and modify some annotation according to our needs.

```{r data}
celPath <- "D:/DataScience/Profile/MultipleMyeloma_GPL570/celFiles"

celList <- list.files(celPath, full.names = T)
celData <- read.celfiles(celList)
rm(celPath, celList)
```

We need to rename the samples according to the expriment design. We will also add some annotation to data, which we will need later for statistical analysis.

```{r addPheno}
colnames(celData)<- c("KMS11.Cfz.1", "KMS11.Cfz.2","KMS11.Cfz.3","KMS34.Cfz.1","KMS34.Cfz.2","KMS34.Cfz.3",
                 "KMS11.1","KMS11.2","KMS11.3","KMS34.1","KMS34.2","KMS34.3")
# phenodata
ph <- celData@phenoData

ph@data[,2] <- c("KMS.11.Cfz","KMS.11.Cfz","KMS.11.Cfz","KMS.34.Cfz","KMS.34.Cfz","KMS.34.Cfz", "KMS.11","KMS.11","KMS.11","KMS.34","KMS.34","KMS.34")
colnames(ph@data)[2] <- "Groups"

ph@data[,3] <- c("KMS11","KMS11","KMS11","KMS34","KMS34","KMS34", "KMS11","KMS11","KMS11","KMS34","KMS34","KMS34")
colnames(ph@data)[3] <- "CellLine"

ph@data[,4] <- c("Cfz", "Cfz", "Cfz", "Cfz", "Cfz", "Cfz", "Ctrl", "Ctrl", "Ctrl", "Ctrl", "Ctrl", "Ctrl")
colnames(ph@data)[4] <- "Treatment"
```

## Quality Control
Before any kind of analysis, we should be sure about the quality of the data. We will apply some quality control tests and visualize them for better understanding of the data.
Principal component analysis will let us know whether samples are separated in an expected way. Samples belonging to one groupe of cell types or treatment should be separated via principal components. 

```{r qualityPCA}
# take log2 of expression and perform a PCA 
exp_raw <- log2(exprs(celData))
PCA_raw <- prcomp(t(exp_raw), sacle. = FALSE)

percentVar <- round(100*PCA_raw$sdev^2/sum(PCA_raw$sdev^2),1)
sd_ratio <- sqrt(percentVar[2]/percentVar[1])

ggData <- data.frame(PC1 = PCA_raw$x[,1],
                     PC2 = PCA_raw$x[,2],
                     Cells = ph@data$CellLine,
                     Treatment = ph@data$Treatment)

ggplot(ggData, aes(PC1, PC2)) + geom_point(aes(shape = Cells, colour = Treatment))+
  ggtitle("PCA of log_transformed raw expression data") + xlab(paste0("PC1, VarExp:", percentVar[1], "%")) +
  ylab(paste0("PC2, VarExp:", percentVar[2], "%")) + theme(plot.title = element_text(hjust = 0.5)) +
  coord_fixed(ratio = sd_ratio) + scale_shape_manual(values = c(4,15)) + scale_color_manual(values = c("darkorange2", "dodgerblue4"))
rm(exp_raw, PCA_raw, percentVar, sd_ratio, ggData)
```

As shown above, two cell types are separated by second principal component, which could be problematic for further analysis. 
To have a closer look at the quality we can draw a boxplot of log intensity of expressions.

```{r qualityBoxplot}
set.seed(521)
oligo::boxplot(celData, target = "core", main = "log2-intensities of raw data")
```

The boxplot shows different median and scale of the boxes, meaning the need of normalization.
Before normalization we perform another quality control test, called relative log expression (RLE). 

```{r RLE}
pData(celData)[,1] <- c("KMS.11.Cfz","KMS.11.Cfz","KMS.11.Cfz","KMS.34.Cfz","KMS.34.Cfz","KMS.34.Cfz", "KMS.11","KMS.11","KMS.11","KMS.34","KMS.34","KMS.34")
pData(celData)[,2] <- c("KMS11","KMS11","KMS11","KMS34","KMS34","KMS34", "KMS11","KMS11","KMS11","KMS34","KMS34","KMS34")
pData(celData)[,3] <- c("Cfz", "Cfz", "Cfz", "Cfz", "Cfz", "Cfz", "Ctrl", "Ctrl", "Ctrl", "Ctrl", "Ctrl", "Ctrl")
names(celData@phenoData@data) <- c("Groups", "Cells", "Treatments")

rleData <- rma(celData, normalize = FALSE) # without normalization to get only log2 values

# The RLE is performed by calculating the median log2 intensity of every transcript across all arrays.
row_medians <- Biobase::rowMedians(as.matrix(exprs(rleData)))

#We then substract this transcript median intensity from every transcript intensity via the sweep function.
RLE_Data <- sweep(exprs(rleData), 1, row_medians)
RLE_Data <- as.data.frame(RLE_Data)

Rle_Data_gathered <- gather(RLE_Data, Treatment, log2_expression_deviation)

ggplot(Rle_Data_gathered, aes(Treatment, log2_expression_deviation)) +
  geom_boxplot(outlier.shape = NA) + ylim(c(-2,2)) + 
  theme(axis.text.x = element_text(colour = "darkgreen", angle = 60, size = 6.5, hjust = 1, face = "bold"))

rm(RLE_Data, Rle_Data_gathered, rleData, row_medians)
```


The boxes with larger extension indicate high deviation from the median. The boxes shifted in y-direction indicate systematically high or lower expression, due to quality issues or batch effects.

## Normalization
Now we will apply RMA algorithm for background correction and normalization.

```{r rma}
eset_norm <- rma(celData)
set.seed(754)
oligo::boxplot(eset_norm, target = "core", main = "log2-intensities of normalized data")
```

The boxplot above shows that RMA algorithm has done a good job in normalizing the data. We can prove it by PCA plot.

```{r normPCA}
norm_exp <- exprs(eset_norm)
PCA <- prcomp(t(norm_exp), scale. = F)
percentVar <- round(100*PCA$sdev^2/sum(PCA$sdev^2),1)
sd_ratio <- sqrt(percentVar[2]/percentVar[1])
ggData <- data.frame(PC1 = PCA$x[,1],
                     PC2 = PCA$x[,2],
                     Cells = pData(eset_norm)$Cells,
                     Treatment = pData(eset_norm)$Treatment)

ggplot(ggData, aes(PC1, PC2)) +
  geom_point(aes(shape = Cells, colour = Treatment)) + 
  ggtitle("PCA plot of normalized data") + 
  xlab(paste0("PC1, VarExp:", percentVar[1], "%")) +
  ylab(paste0("PC2, VarExp:", percentVar[2], "%")) +
  theme(plot.title = element_text(hjust = 0.5)) +
  coord_fixed(ratio = sd_ratio) + 
  scale_shape_manual(values = c(4,15)) +
  scale_color_manual(values = c("darkorange2", "dodgerblue4"))
rm(norm_exp, PCA, percentVar, sd_ratio, ggData)
```

After normalization, the cell types are separated by first principal component. 


## Filtering
Next, we will exclude the lowly expressed genes. To this aim, we will calculate row-wise medians of gene expression intensity. Then we will set a cutoff to to remove the defined expressions.

```{r histFilter}
eset_medians <- rowMedians(exprs(eset_norm))
hist(eset_medians, 100, col = "cornsilk1", freq = FALSE,
     main = "Histogram of the median intensities",
     border = "antiquewhite4", xlab = "Median intensities")
# set cutoff threshold left to the histogram peak
cut_thereshold <- 4
abline(v=cut_thereshold, col = "coral4", lwd = 2)
```

We can exlude the low medians, which are enriched on the left hand side of the histogram.


```{r filtering}
idx_cut_threshold <- apply(exprs(eset_norm), 1, function(x){
  sum(x>cut_thereshold) >= 3}) # number of arrays where the median intesity passes the threshold is greater than the sample cutoff(we have 3 samples per replicate)

table(idx_cut_threshold) # 3437 arrays can be removed

eset_norm_filtered <- subset(eset_norm, idx_cut_threshold)

rm(cut_thereshold, eset_medians, idx_cut_threshold)
```

To perform differntial expression analysis, we need first to add some more annotation information to featureData of our expression set.


```{r annotation}
anno_eset <- AnnotationDbi::select(hgu133plus2.db, keys = (featureNames(eset_norm_filtered)),
                                   columns = c("SYMBOL", "GENENAME"),
                                   keytype = "PROBEID")
anno_eset <- subset(anno_eset, !is.na(SYMBOL)) # filterout probes that do not map to a gene
```

Many transcript identifiers are mapped to multiple gene symbols, which we are going to remove.
First we filter the Probe Ids with multiple matches. Then we subset the expression set based on filtered probe IDs. We will also remove the corresponding IDs from annotation set. Finaly we will merge the expression set and annotation set to get a final set to work with.

```{r multMapping}
anno_grouped <- group_by(anno_eset, PROBEID)
anno_summarized <- summarize(anno_grouped, no_of_matches = n_distinct(SYMBOL))


anno_filtered <- filter(anno_summarized, no_of_matches > 1) # filter for PROBEIDs with multiple matches
head(anno_filtered)
nrow(anno_filtered) # 2157 PROBEID with multiple matches, difficult to decide which mapping is correct. Therefoe we exclude these clusters


ids_to_exclude <- (featureNames(eset_norm_filtered) %in% anno_filtered$PROBEID)
table(ids_to_exclude)

# subset to final eset
eset_final <- subset(eset_norm_filtered, !ids_to_exclude)
validObject(eset_final)# TRUE

# we should also remove probe IDs from anno_eset 
head(anno_eset)

# first we store feature data in eset_final, defining PROBEID column
fData(eset_final)$PROBEID <- rownames(fData(eset_final))

# left_join anno_eset
fData(eset_final) <- left_join(fData(eset_final), anno_eset)

# restore row names
rownames(fData(eset_final)) <- fData(eset_final)$PROBEID
validObject(eset_final)

rm(anno_eset, anno_filtered, anno_grouped, anno_summarized, eset_norm, eset_norm_filtered, ids_to_exclude)
```

## Differential Gene Expression
As we see in exprimental design, two cell lines have been treated with anti cancer drug. Therefor we should consider two factors affecting the outcomes:
1. Cell types: KMS11 and KMS34
2. Treatment: control vs. Carfilzomib

One can do a joint model of differential expression analysis, but the result could be affected by two factors. As we could see before the main difference in expression pattern is coming from the type of the cells.

We will create two design matrices, one for each cell line. Then we will explore the differential gene expression separately for each cell line.

```{r DE}
Groups <- as.character(pData(eset_final)$Groups)
Cells <- as.character(pData(eset_final)$Cells)
Treatment <- as.character(pData(eset_final)$Treatment)

# according to the original paper we look for the changes in transcription occur between treatments. we create two designe matrices one for each cell line

i_11 <- Groups[Cells == "KMS11"]
design_eset_11 <- model.matrix(~ 0 + Treatment[Cells == "KMS11"] + i_11)
colnames(design_eset_11)[1:2] <- c("Cfz", "Ctrl")
rownames(design_eset_11) <- i_11



i_34 <-  Groups[Cells == "KMS34"]
design_eset_34 <- model.matrix(~ 0 + Treatment[Cells == "KMS34"] + i_34)
colnames(design_eset_34)[1:2] <- c("Cfz", "Ctrl")
rownames(design_eset_34) <- i_34




contrast_matrix_11 <- makeContrasts(Cfz-Ctrl, levels = design_eset_11)
eset_fit_11 <- eBayes(contrasts.fit(lmFit(eset_final[,Cells =="KMS11"],
                                          design = design_eset_11),
                                    contrast_matrix_11))

volcano_names <- ifelse(abs(eset_fit_11$coefficients)>=1,
                        eset_fit_11$genes$SYMBOL, NA)
volcanoplot(eset_fit_11, style = "p-value", highlight = 10, names = volcano_names, main = "KMS11 Cfs vs. Ctrl",
            xlab = "Log2 Fold Change", ylab = NULL, pch = 16, cex = 0.35)
```

```{r DE2}

contrast_matrix_34 <- makeContrasts(Cfz-Ctrl, levels = design_eset_34)
eset_fit_34 <- eBayes(contrasts.fit(lmFit(eset_final[,Cells =="KMS34"],
                                          design = design_eset_34),
                                    contrast_matrix_34))

volcano_names <- ifelse(abs(eset_fit_34$coefficients)>=1,
                        eset_fit_34$genes$SYMBOL, NA)
volcanoplot(eset_fit_34, style = "p-value", highlight = 10, names = volcano_names,main = "KMS34 Cfz vs. Ctrl",
            xlab = "Log2 Fold Change", ylab = NULL, pch = 16, cex = 0.35)

rm(design_eset_11, design_eset_34, contrast_matrix_11, contrast_matrix_34, volcano_names)


```

we will set another cutoff, this time for genes with FDR < 5%, and select the top 50 genes. Then we will look for the ones that are listed in both cell lines.

```{r filterFDR}
table_11 <- topTable(eset_fit_11, number = Inf)
table_11 <- table_11[order(table_11$adj.P.Val),]

selected_11 <- subset(table_11, adj.P.Val < 0.05)$PROBEID
eset_11 <- eset_final[selected_11,]
table_11[1:20,]
```


```{r FDR34}
table_34 <- topTable(eset_fit_34, number = Inf)
table_34 <- table_34[order(table_34$adj.P.Val),]

selected_34 <- subset(table_34, adj.P.Val < 0.05)$PROBEID
eset_34 <- eset_final[selected_34,]
table_34[1:20,]

shared_genes <- selected_11[c(which(selected_11 %in% selected_34))]
shared_DE <-AnnotationDbi::select(hgu133plus2.db, keys = (shared_genes),
                      columns = c("SYMBOL", "GENENAME"),
                      keytype = "PROBEID")
dim(shared_DE)
shared_DE[1:10,]
```

189 genes are differentially expressed under Carfilzomib treatmnet (aquired resistance) in both cell lines. Top 10 of them are printed above.

Now, we will draw a heatmap for differentially expressed gene in each set.

```{r heatmap11}
heat_table <- table_11[1:50,]
heat_ID <- heat_table$PROBEID
heat_set <- eset_final[heat_ID,]
exprs_11 <- exprs(heat_set)
row.names(exprs_11) <- heat_table$SYMBOL


pheatmap(exprs_11,
         annotation_col = heat_set@phenoData@data[,c("Cells", "Treatments")],
         scale = "row",
         cutree_rows = 4, cutree_cols = 4,
         main = "Clustering heatmap of KMS11 Cfz vs. Ctrl")
rm(exprs_11)
```

```{r heatmap34}
heat_table <- table_34[1:50,]
heat_ID <- heat_table$PROBEID
heat_set <- eset_final[heat_ID,]
exprs_34 <- exprs(heat_set)
row.names(exprs_34) <- heat_table$SYMBOL


pheatmap(exprs_34,
         annotation_col = heat_set@phenoData@data[,c("Cells", "Treatments")],
         scale = "row",
         cutree_rows = 4, cutree_cols = 4,
         main = "Clustering heatmap of KMS34 Cfz vs. Ctrl")
rm(heat_table, heat_ID, heat_set, exprs_34)
```

## GO Enrichment Analysis
Now we can do some enrichment analysis of DE genes, base on gene ontology. 
First we find a set of background genes which are similar in expression to DE genes, and look at their distribution to avoid the biased background.

```{r backGenes}
back_genes_id11 <- genefinder(eset_final, as.character(selected_11), method = "manhattan", scale = "none")
back_genes_id11 <- sapply(back_genes_id11, function(x)x$indices)

back_genes_id34 <- genefinder(eset_final, as.character(selected_34), method = "manhattan", scale = "none")
back_genes_id34 <- sapply(back_genes_id34, function(x)x$indices)


back_genes11 <- featureNames(eset_final)[back_genes_id11]
back_genes11 <- setdiff(back_genes11, selected_11)
intersect(back_genes11, selected_11)
length(back_genes11)


back_genes34 <- featureNames(eset_final)[back_genes_id34]
back_genes34 <- setdiff(back_genes34, selected_34)
intersect(back_genes34, selected_34)
length(back_genes34)
```

The intersect = 0 confirmes that the back_genes are not included in selected DE genes. 
To check whether the background genes are not biased for the gene enrichment analysis, we will creat a multidensity plot, where the mean expression is plotted on the x-axis, and check for the similarity of the distribution of back-genes and fore-genes.

```{r multiDensity_11}
multidensity(list(
  all = table_11[,"AveExpr"],
  fore = table_11[selected_11, "AveExpr"],
  back = table_11[rownames(table_11) %in% back_genes11, "AveExpr"]),
  col = c("#e46981", "#ae7ee2", "#a7ad4a"),
  xlab = "mean expression",
  main = "DE genes for KMS11-background-matching")
```

```{r multiDensity_34}
multidensity(list(
  all = table_34[,"AveExpr"],
  fore = table_34[selected_34, "AveExpr"],
  back = table_34[rownames(table_34) %in% back_genes34, "AveExpr"]),
  col = c("#e46981", "#ae7ee2", "#a7ad4a"),
  xlab = "mean expression",
  main = "DE genes for KMS11-background-matching")
```

The distribution of bak and fore-genes are similar in both sets. Now we can perform GO enrichment analysis.


```{r topGO}
gene_ID11 <- rownames(table_11)
in_universe <- gene_ID11 %in% c(selected_11, back_genes11)
in_selection <- gene_ID11 %in% selected_11

all_genes11 <- in_selection[in_universe]
all_genes11 <- factor(as.integer(in_selection[in_universe]))
names(all_genes11) <- gene_ID11[in_universe]



gene_ID34 <- rownames(table_34)
in_universe <- gene_ID34 %in% c(selected_34, back_genes34)
in_selection <- gene_ID34 %in% selected_34

all_genes34 <- in_selection[in_universe]
all_genes34 <- factor(as.integer(in_selection[in_universe]))
names(all_genes34) <- gene_ID34[in_universe]

## initialize topGO data set
top_GO_data_11 <- new("topGOdata", ontology = "BP", allGenes = all_genes11,
                      nodeSize = 10, annot = annFUN.db, affyLib = "hgu133plus2.db") # nodeSize=10 specifies a minimum size of GO category, categories with less than 10 genes are not included


top_GO_data_34 <- new("topGOdata", ontology = "BP", allGenes = all_genes34,
                      nodeSize = 10, annot = annFUN.db, affyLib = "hgu133plus2.db")


### testing with Fisher classic and elim algorithm
result11_topGO_elim <- runTest(top_GO_data_11, algorithm = "elim", statistic = "Fisher")
result11_topGO_classic <- runTest(top_GO_data_11, algorithm = "classic", statistic = "Fisher")



result34_topGO_elim <- runTest(top_GO_data_34, algorithm = "elim", statistic = "Fisher")
result34_topGO_classic <- runTest(top_GO_data_34, algorithm = "classic", statistic = "Fisher")

res_top_GO_11 <- GenTable(top_GO_data_11, Fisher.elim = result11_topGO_elim, Fisher.classic = result11_topGO_classic,
                          orderBy = "Fisher.elim", topNodes = 20)
genes_top_GO_11 <- printGenes(top_GO_data_11, whichTerms = res_top_GO_11$GO.ID, chip = "hgu133plus2.db", geneCutOff = 1000)

res_top_GO_11$sig_genes <- sapply(genes_top_GO_11, function(x){
  str_c(paste0(x[x$"raw p-value" == 2, "Symbol.id"], ";"), collapse = "")
})


res_top_GO_34 <- GenTable(top_GO_data_34, Fisher.elim = result34_topGO_elim, Fisher.classic = result34_topGO_classic,
                          orderBy = "Fisher.elim", topNodes = 20)
genes_top_GO_34 <- printGenes(top_GO_data_34, whichTerms = res_top_GO_34$GO.ID, chip = "hgu133plus2.db", geneCutOff = 1000)

res_top_GO_34$sig_genes <- sapply(genes_top_GO_34, function(x){
  str_c(paste0(x[x$"raw p-value" == 2, "Symbol.id"], ";"), collapse = "")
})


```

Now we can look at top terms of GO analysis
```{r GOplot}
res_top_GO_11 %>% 
     top_n(10, wt=1/as.numeric(Fisher.elim)) %>% 
     mutate(hitsPerc=Significant*100/Annotated) %>% 
     ggplot(aes(x=hitsPerc, 
                y=Term, 
                colour=Fisher.elim, 
                size=Significant)) +
     geom_point() +
     expand_limits(x=0) +
     labs(x="Hits (%)", y="GO term", colour="p value", size="Count") + ggtitle("GO terms for KMS11")
```
In KMS11, regulation of cellular amino acid and negative regulation of canonical Wnt signaling are at top.

```{r GOplot34}
res_top_GO_34 %>% 
     top_n(10, wt=1/as.numeric(Fisher.elim)) %>% 
     mutate(hitsPerc=Significant*100/Annotated) %>% 
     ggplot(aes(x=hitsPerc, 
                y=Term, 
                colour=Fisher.elim, 
                size=Significant)) +
     geom_point() +
     expand_limits(x=0) +
     labs(x="Hits (%)", y="GO term", colour="p value", size="Count") + ggtitle("GO terms for KMS34")
```
In KMS34, negative regulation of IL6 and regulation of cellular response to IF gamma are at top.

## Pathway enrichment analysis using reactom

 
